---
title: "Randomisation of samples on genotyping chip"
subtitle: "Randomise samples to avoid plate / batch effects in analyses"
author: "Anna Vinkhuyzen"
email: "hacqconsulting@gmail.com"
date: "18/08/2020"
output: html_document
---

***
***

```{r packages, message=FALSE, echo=FALSE}
library(rmarkdown)  
library(qqman)
library(plyr)
library(dplyr)
library(platetools)
library(ggplot2)
library(viridis)
library(data.table)
library(janitor)
library(fabricatr) 
library(crosstabr) # devtools::install_github("tklebel/crosstabr")

```

# Brief

*From lab manager:*
 
We have a target list with Autism cases and healthy controls. We wish for you to generate a plating manifest for running on GSA chips (Infinium Global Screening / Genome-Wide Association, GWA). Samples should be randomly distributed across plates and visual inspection is required to make sure no accidental grouping of cases / sex / cohorts excist on particular plates.

Please could you leave **D3**  free for a technical control.

The attached data contains all the patient information, below is the list of column headers in the file. Hoping that you will have all the info you need to generate a manifest.
 
* *ID*, unique study ID
* *case.control*, control = 0, case = 1
* *sex*, male = 0, female = 1
* *cohort*, 5 participating cohorts, cohort.1 - cohort.5

***
***

```{r data, message=FALSE, warning=FALSE, echo=FALSE, results=FALSE}

## Fabricate data for presentation purposes

n.plates <- 10 # number of plates
n.wells <- n.plates*96
n.technical.controls <- n.plates
n.participants <- n.wells - n.technical.controls

## variables required: 
  # ID
  # Autism disease status (case.control)
  # sex
  # cohort

## create correlated vectors with 
  # Autism cases 
  # Cohorts
  # Sex

ID <- round(runif(n.participants, min=10000, max=99999),0)

### 950 study participants
## approx 60% case
## approx 50% female
## unequal distribution across cohorts
## both sex and cohort are associated with proportion of cases, 
  # i.e, autism more common in males and more common in some cohorts

data <- fabricate(
  N = 950,
  case.control = draw_binomial(prob = 0.6, trials = 1, N = N),
  sex = correlate(given = case.control,
                                    rho = 0.7,
                                    draw_binomial,
                                    prob = 0.4,
                                    trials = 1),
  cohort = correlate(given = case.control,
                             rho = 0.3,
                             draw_binomial,
                             prob = 0.5,
                             trials = 4)
)

data$cohort <- replace(data$cohort, data$cohort==0, 5)

```

## Descriptive analyses of the provided data

```{r visual inspection, message=FALSE, warning=FALSE, echo=FALSE}

## create temp data.frame with informative labels
data.tmp <- data

data.tmp$sex <- replace(data.tmp$sex, data.tmp$sex==0, "male")
data.tmp$sex <- replace(data.tmp$sex, data.tmp$sex==1, "female")

data.tmp$case.control <- replace(data.tmp$case.control, data.tmp$case.control==0, "control")
data.tmp$case.control <- replace(data.tmp$case.control, data.tmp$case.control==1, "case")

# bar plot case.control & cohort
counts <- table(data.tmp$case.control, data.tmp$cohort)
barplot(counts, main="case.control by cohort",
  xlab="cohort", ylab="number of cases/controls", col=c("darkblue","red"),
  legend = c("control", "case")) 

# bar plot case.control & sex
counts <- table(data.tmp$case.control, data.tmp$sex)
barplot(counts, main="case.control by sex",
  xlab="sex", ylab="number of cases/controls", col=c("darkblue","red"),
  legend = c("control", "case")) 

# cross-table case.control & sex
data.tmp %>%
  crosstab(case.control~sex)

# cross-table case.control & cohort
data.tmp %>%
  crosstab(case.control~cohort)

```


## Creating plates

I am assuming plates with  96 wells, that is 8 rows (A,B,C,D,E,F,G) and 12 columns (1:12)

I include 1 technical control on each plate: well D3 

Samples will be randomized across plates

```{r create plates, message=FALSE, warning=FALSE, echo=FALSE, results=FALSE }

## create manifests data.frames

n.plates ## based on the above
n.rows <- 8 ## fixed
n.cols <- 12 ## fixed

# plates
rows.per.plate <- rep(c("A", "B", "C", "D", "E","F","G","H"),n.cols)
cols.per.plate <- rep(1:n.cols, each=n.rows)

rows.all.plates <- rep(rows.per.plate, n.plates)
cols.all.plates <- rep(cols.per.plate, n.plates)

plate.number <- rep(1:n.plates, each=96)

plate.design <- data.frame(plate.number,rows.all.plates,cols.all.plates) 
names(plate.design) [2] <- "row"
names(plate.design) [3] <- "column"

#split the manifest data.frame into two data.frames, 1 for control and 1 for samples 

## we have 1 technical control on each plate: D3
plate.design.controls <- plate.design[which(plate.design$row=="D" & 
                                        plate.design$column==3),]
## we have n.participants number of samples
plate.design.samples <- plate.design[!(plate.design$row=="D" & plate.design$column==3),]
## attach seq.number to enable merging with data
plate.design.samples$seq.number <- seq(1:n.participants)

```


```{r randomize, message=FALSE, warning=FALSE, echo=FALSE, results=FALSE}

# randomize input set and add to plates

## randomize set
head(data)
dim(data)
## attach seq.number to enable merging with plate.design
data$seq.number <- sample(1:nrow(data), replace=FALSE) 
data.2 <- data[order(data$seq.number),]

####################################################################################

## combine manifest with data
plate.design.samples <- merge(plate.design.samples, data.2, all = TRUE)

## add  controls
plate.design.controls[,colnames(data.2)] <- "TECH_CONTROL"

plate.design.all <- rbind(plate.design.samples, plate.design.controls)
plate.design.all <- plate.design.all[order(plate.design.all$plate.number, plate.design.all$row, plate.design.all$column),]
plate.design.all

```


## Plots of the plates with randomized samples

Note that variables are displayed as numerical variables as the plotting code requires variables to be numeric. 

```{r, message=FALSE, warning=FALSE, echo=FALSE, fig.width=15, fig.height=30}
names <- c("case.control", "sex","cohort", "plate.number","row","column")
plate.design.for.plot <- plate.design.all[names]
#head(plates.for.plot)

## case.control
plate.design.for.plot$case.control <- replace(plate.design.for.plot$case.control, plate.design.for.plot$case.control=="TECH_CONTROL", 2)
plate.design.for.plot$case.control <- as.numeric(plate.design.for.plot$case.control)

# cohort
plate.design.for.plot$cohort <- replace(plate.design.for.plot$cohort, plate.design.for.plot$cohort=="TECH_CONTROL", 6)
plate.design.for.plot$cohort <- as.numeric(plate.design.for.plot$cohort)

# sex
plate.design.for.plot$sex <- replace(plate.design.for.plot$sex, plate.design.for.plot$sex=="TECH_CONTROL", 2)
plate.design.for.plot$sex <- as.numeric(plate.design.for.plot$sex)

names(plate.design.for.plot)[names(plate.design.for.plot) == 'plate.number'] <- 'Sample_Plate'

plate.design.for.plot$Sample_Well <- paste(plate.design.for.plot$row, plate.design.for.plot$column, sep="")

## case.control
 raw_grid(data = plate.design.for.plot$case.control,
         well = plate.design.for.plot$Sample_Well,
         plate_id = plate.design.for.plot$Sample_Plate       ) +
  scale_fill_distiller(type = "div") +
  ggtitle("Plates with case-control status; 0=control; 1=case, 2=technical control")

 ## sex
 raw_grid(data = plate.design.for.plot$sex,
         well = plate.design.for.plot$Sample_Well,
         plate_id = plate.design.for.plot$Sample_Plate       ) +
  scale_fill_distiller(type = "div") +
  ggtitle("Plates with sex status; 0=male; 1=female; 2=technical control")
 
  ## cohort
 raw_grid(data = plate.design.for.plot$cohort,
         well = plate.design.for.plot$Sample_Well,
         plate_id = plate.design.for.plot$Sample_Plate       ) +
  scale_fill_distiller(type = "div") +
  ggtitle("Plates with cohort status: 1=cohort.1; 2=cohort.2; 3=cohort.3; 4=cohort.4; 5=cohort.5; 6=technical control")
 
 #dev.off()

```




```{r, message=FALSE, warning=FALSE, echo=TRUE}
# write manifest

## write file for lab
fileout <- "~/Documents/Prive/HACQ/Presentations/20200818_RLadies/20200818_Autism_GSA_Manifest.csv" 
write.csv(plate.design.all,fileout, row.names=FALSE, quote=FALSE, na="NA")

```
