---
title: "R-ladies 21.08.2019"
author: "Gunn-Helen Moen"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Quick introduction to Rmarkdown
You write your document in plain text and you can include *italics* and **bold**
formats in addition to superscript^2^ and [links](www.rstudio.com). When you
want a new paragraph you end the line with two spaces.
You can insert a header by using one or more hashtags:

# Header 1

## Header 2

### and so on

>block quotes

* unordered lists
* with several items
+ and sub-items
+ more sub-items

1. Ordered lists
2. Like this
+ with adidional sub-items

You can also make tables:

Table header 1 | Table header 2
---------------| --------------
Cell 1 | Cell 2
Cell 3 | Cell 4
To add a code chunck to your document write three back ticks followed by {r}. To
end the code chunck add three more back ticks.

```{r}
paste("Hello", "R-ladies!")
```

Within the braces you can add chuck options, `echo=FALSE` will prevent the
source code from being displayed and the code will look like this:
```{r echo=FALSE}
paste("Hello", "R-ladies!")
```
Another usefull one is `eval=TRUE` for running the code in the code chunk and
`eval=FALSE` for not running the code in the code chunk.
To add comments to your code just use # withing the code chuck window
```{r eval=FALSE}
paste("Hello", "R-ladies!") #Add comment
```
*Here I've used `eval=FALSE` so that the code won't start running.
---
- [Rmarkdown cheetsheet is a useful
resource](https://www.rstudio.com/wp-content/uploads/2016/03/rmarkdown-cheatshee
t-2.0.pdf)
- [As well as the Rmarkdown
reference](https://www.rstudio.com/wp-content/uploads/2015/03/rmarkdown-referenc
e.pdf)
---
# ggplot2
GGplot2 is one of the graphing library's in R.
```{r}
#install.packages("ggplot2")
library("ggplot2")
```
### ggplot2 graphing consists of three parts
1. **aesthetic**
+ An aesthetic that tells ggplot which variables are being mapped to the x axis,
y axis, (and often other attributes of the graph, such as the color fill).
Intuitively, the aesthetic can be thought of as what you are graphing.
2. **geom**
+ A geom or geometry that tells ggplot about the basic structure of the graph.
Intuitively, the geom can be thought of as how you are graphing it.
3. **options**
+ Other options, such as a graph title, axis labels and overall theme for the
graph.
---

##### For this demonstration I will use the R example data mtcars.
Loading the data and having a look:
```{r}
data(mtcars)
head(mtcars)
```
---
The structure of building a ggplot consist of a `aesthetic` + `one or more
geoms` + `optional elements`.

```{r}
ggplot(mtcars, #my dataset
aes(x=mpg)) + #what I am graphing - here miles per gallon from mtcars dataset
geom_histogram(fill="blue", #how is it being graphed
color="black",
bins = 20)
```

From here you can start adding labels
```{r}
ggplot(mtcars, #my dataset
aes(x=mpg)) + #what I am graphing - here miles per gallon from mtcars dataset
geom_histogram(fill="blue", #how is it being graphed
color="black",
bins = 20) +
labs(title = "Miles per gallon",
subtitle = "Subtitle",
caption = "mtcars dataset",
x = "mpg",
y = "count")
```
##### Assigning the aesthetic to an plot
You can assign your basic ggplot function to a plot so that you wont have to
call it everytime. For instance:
```{r}
p<- ggplot(mtcars, aes(x=mpg))
```
#### The different geoms
There are lot of different geoms you can use.
##### For one variable you can use plots like:
+ `+ geom_dotplot()` To add a dotplot geometry to the graph.
+ `+ geom_histogram()` To add a histogram geometry to the graph.
+ `+ geom_density()` To add a density plot to the graph.
+ `+ geom_qq()` To add a QQ plot ot the graph
Example of plot:
```{r}
p + geom_dotplot()
```
##### For two variable you can use plots like:
+ `+ geom_point()` To add a point (scatterplot) geometry to the graph.
+ `+ geom_smooth()` To add a smoother to the graph.
+ `+ geom_boxplot()` To add a boxplot to the graph.
To vizualise we need a aesthetic with two variables
```{r}
p2<- ggplot(mtcars, aes(x=mpg, y=hp)) 
#two continous variables (miles pergallon and horsepower)
```
Examples of plots:
Scatterplot:
```{r}
p2 + geom_point()
```
Scatterplot with Smoother:
```{r}
p2 + geom_smooth()
```
Combining the two:
```{r}
p2 + geom_smooth() + geom_point()
```

#### Saving your plot
Save your plot by using `ggsave("plot.png", width = 5, height = 5)`

[ggplot2 cheetsheet is a usefull
resourse](https://github.com/rstudio/cheatsheets/blob/master/data-visualization-
2.1.pdf)
