---
title: "R-Ladies' - Data Manipulation with R"
output: html_notebook
---
Speaker: Ruth Chirinos


# Data transformation
## Prerequisites
```{r}
#install.packages("dplyr")
library(tidyverse)
library(nycflights13)
```

## Data set nycflights13 
```{r}
flights
#
?flights
```

More ways to learn basic info on a data.frame.
```{r}
names(flights)
dim(flights)    # ?dim dimension
ncol(flights)   # ?ncol number of columns
nrow(flights)   # ?nrow number of rows
```

```{r}
summary(flights)
#install.packages('skimr')
library(skimr) 
skim(flights)
```

Ver el tibble (data frame)
```{r}
#View(flights)
glimpse(flights)
```

## dplyr basics
- Pick observations by their values (filter()).
- Reorder the rows (arrange()).
- Pick variables by their names (select()).
- Create new variables with functions of existing variables (mutate()).
- Collapse many values down to a single summary (summarise()).

### Filter rows with filter()

```{r}
filter(flights, month == 1, day == 1)
```

dplyr functions never modify their inputs, so if you want to save the result, you’ll need to use the assignment operator, <-
```{r}
( jan1 <- filter(flights, month == 1, day == 1) )

```
### Comparisons


```{r}
# filter(flights, month = 1)
```
```{r}
sqrt(2) ^ 2 == 2
1 / 49 * 49 == 1
#
near(sqrt(2) ^ 2,  2)
near(1 / 49 * 49, 1)
```
### Logical operators


Boolean operators: & is “and”, | is “or”, and ! is “no”
```{r}
filter(flights, month == 11 | month == 12)
```
Example with %in%
```{r}
( nov_dec <- filter(flights, month %in% c(11, 12)) )
```
If you wanted to find flights that weren’t delayed (on arrival or departure) by more than two hours, you could use either of the following two filters:

```{r}
filter(flights, !(arr_delay > 120 | dep_delay > 120))
filter(flights, arr_delay <= 120, dep_delay <= 120)
```

### Missing values

```{r}
NA > 5
#> [1] NA
10 == NA
#> [1] NA
NA + 10
#> [1] NA
NA / 2
#> [1] NA
```

```{r}
NA == NA
#> [1] NA
```
```{r}
x <- NA
y <- NA
x == y
#> [1] NA
# We don't know!
```
```{r}
is.na(x)
```
```{r}
df <- tibble(x = c(1, NA, 3))
df
filter(df, x > 1)
filter(df, is.na(x) | x > 1)
```

## Arrange rows with arrange()

```{r}
arrange(flights, year, month, day)
```
```{r}
arrange(flights, desc(dep_delay))
```
Missing Values
```{r}
df <- tibble(x = c(5, 2, NA))
arrange(df, x)
arrange(df, desc(x))
```
## Select columns with select()

```{r}
select(flights, year, month, day)
```
```{r}
select(flights, year:day)
```
```{r}
select(flights, -(year:day))
```
```{r}
rename(flights, tail_num = tailnum)
```
```{r}
select(flights, time_hour, air_time, everything())
```

## Add new variables with mutate()

```{r}
flights_sml <- select(flights, 
  year:day, 
  ends_with("delay"), 
  distance, 
  air_time
)
mutate(flights_sml,
  gain = dep_delay - arr_delay,
  speed = distance / air_time * 60
)
```

Note that you can refer to columns that you’ve just created:
```{r}
mutate(flights_sml,
  gain = dep_delay - arr_delay,
  hours = air_time / 60,
  gain_per_hour = gain / hours
)
```

If you only want to keep the new variables, use transmute():
```{r}
transmute(flights,
  gain = dep_delay - arr_delay,
  hours = air_time / 60,
  gain_per_hour = gain / hours
)
```
## Grouped summaries with summarise()

### Group By
```{r}
(by_day <- group_by(flights, year, month, day))
```


```{r}
summarise(flights, delay = mean(dep_delay, na.rm = TRUE))
```
```{r}
(by_day <- group_by(flights, year, month, day))
summarise(by_day, delay = mean(dep_delay, na.rm = TRUE))
```

### Combining multiple operations with the pipe

```{r}
delay <- flights %>% 
  group_by(dest) %>% 
  summarise(
    count = n(),
    dist = mean(distance, na.rm = TRUE),
    delay = mean(arr_delay, na.rm = TRUE)
  ) %>% 
  filter(count > 20, dest != "HNL")
#
ggplot(data = delay, mapping = aes(x = dist, y = delay)) +
  geom_point(aes(size = count), alpha = 1/3) +
  geom_smooth(se = FALSE)
```
### Missing values
```{r}
flights %>% 
  group_by(year, month, day) %>% 
  summarise(mean = mean(dep_delay, na.rm = TRUE))
```

```{r}
not_cancelled <- flights %>% 
  filter(!is.na(dep_delay), !is.na(arr_delay))
#
not_cancelled %>% 
  group_by(year, month, day) %>% 
  summarise(mean = mean(dep_delay))
```

The story is actually a little more nuanced. We can get more insight if we draw a scatterplot of number of flights vs. average delay:
```{r}
delays <- not_cancelled %>% 
  group_by(tailnum) %>% 
  summarise(
    delay = mean(arr_delay, na.rm = TRUE),
    n = n()
  )

ggplot(data = delays, mapping = aes(x = n, y = delay)) + 
  geom_point(alpha = 1/10)
```

With a filter n > 25
```{r}
delays %>% 
  filter(n > 25) %>% 
  ggplot(mapping = aes(x = n, y = delay)) + 
    geom_point(alpha = 1/10) # +
  #geom_smooth(se = FALSE)
```

### Useful summary functions

Measures of location: we’ve used mean(x), but median(x) is also useful.
```{r}
not_cancelled %>% 
  group_by(year, month, day) %>% 
  summarise(
    avg_delay1 = mean(arr_delay),
    avg_delay2 = mean(arr_delay[arr_delay > 0]) # the average positive delay
  )
```

Measures of spread: sd(x), IQR(x), mad(x)
```{r}
not_cancelled %>% 
  group_by(dest) %>% 
  summarise(distance_sd = sd(distance)) %>% 
  arrange(desc(distance_sd))
```

Measures of rank: min(x), quantile(x, 0.25), max(x)
```{r}
# When do the first and last flights leave each day?
not_cancelled %>% 
  group_by(year, month, day) %>% 
  summarise(
    first = min(dep_time),
    last = max(dep_time)
  )
```
Measures of position: first(x), nth(x, 2), last(x)
```{r}
not_cancelled %>% 
  group_by(year, month, day) %>% 
  summarise(
    first_dep = first(dep_time), 
    last_dep = last(dep_time)
  )
```

Filtering on Ranks
```{r}
not_cancelled %>% 
  group_by(year, month, day) %>% 
  mutate(r = min_rank(desc(dep_time))) %>% 
  filter(r %in% range(r))
```

Counts: You’ve seen n(),
```{r}
# Which destinations have the most carriers?
not_cancelled %>% 
  group_by(dest) %>% 
  summarise(carriers = n_distinct(carrier)) %>% 
  arrange(desc(carriers))
```

Counts are so useful that dplyr provides a simple helper if all you want is a count:
```{r}
not_cancelled %>% 
  count(dest)
```

Counts and proportions of logical values: sum(x > 10), mean(y == 0)
```{r}
# How many flights left before 5am? (these usually indicate delayed
# flights from the previous day)
not_cancelled %>% 
  group_by(year, month, day) %>% 
  summarise(n_early = sum(dep_time < 500))

# What proportion of flights are delayed by more than an hour?
not_cancelled %>% 
  group_by(year, month, day) %>% 
  summarise(hour_prop = mean(arr_delay > 60))
```

### Grouping by multiple variables
When you group by multiple variables, each summary peels off one level of the grouping. That makes it easy to progressively roll up a dataset:
```{r}
daily <- group_by(flights, year, month, day)
(per_day   <- summarise(daily, flights = n()))
#
(per_month <- summarise(per_day, flights = sum(flights)))
#
(per_year  <- summarise(per_month, flights = sum(flights)))

```


### Ungrouping
```{r}
daily
#
daily %>% 
  ungroup() %>%             # no longer grouped by date
  summarise(flights = n())  # all flights
```

### Grouped mutates (and filters)
- Find the worst members of each group:
```{r}
flights_sml %>% 
  group_by(year, month, day) %>%
  filter(rank(desc(arr_delay)) < 10)
```

- Find all groups bigger than a threshold:
```{r}
popular_dests <- flights %>% 
  group_by(dest) %>% 
  filter(n() > 365)
#
popular_dests
```

- Standardise to compute per group metrics:
```{r}
popular_dests %>% 
  filter(arr_delay > 0) %>% 
  mutate(prop_delay = arr_delay / sum(arr_delay)) %>% 
    select(year:day, dest, arr_delay, prop_delay)
```



