---
title: "Tips n Tricks"
author: "Paula Andrea Martinez"
date: "`r Sys.Date()`"
output:
  tufte::tufte_html: default
  tufte::tufte_handout:
    citation_package: natbib
    latex_engine: xelatex
  tufte::tufte_book:
    citation_package: natbib
    latex_engine: xelatex
bibliography: skeleton.bib
link-citations: yes
---

```{r setup, include=FALSE}
library(tufte)
# invalidate cache when the tufte version changes
knitr::opts_chunk$set(tidy = FALSE, cache.extra = packageVersion('tufte'))
options(htmltools.dir.version = FALSE)
```

`r newthought('For beginners')`

# *Manage your project structure

## Tip - Create an RStudio Project
`File > New Project`

## Tip - Create a folder structure for your project
You can click on `New Folder` and create your structure
or you can do it from a script
```{r eval=FALSE}
dir.create("data")
```

## Trick - Create a package that does that for you
tidyproject ^[Find it on GitHub orchid00/tidyproject]

```{r eval=FALSE}
remotes::install_github("orchid00/tidyproject")
library(tidyproject)
# This function created my folder structure
# with data, scripts, plots, rmarkdown
#createStr()
```

## Tip - Use here from the here package
It will save you headaches about paths you don't need to worry about

```{r eval=FALSE}
install.packages("here")
here("data","myfile.csv")
```

## Trick - Open Recent
Please don’t spend time navigating through your folder structure to find a 
recent project or file.

*File > Recent Files*
*File > Recent Projects*

Also, *top right corner arrow down*

## Trick - Find in Files
*Shortcut* `Cmd / Ctrl + Shift + F`

## Tip - Add comments
Use a # at the beginning of the line
or at the end of the line

## Trick - Add comments
*Shortcut* `Cmd / Ctrl + Shift + C`

```{r}
# comment
```


## Tip - Add sections to your code
At the end of a section add
4 dashes - or 4 hashes #
```{r}
# this is a section ----
```

## Tip - Get data in
You can download files from the web
directly into your project
```{r eval=FALSE}
?download.file
```

# *Cheatsheets

## Tip - used them when you are learning or to refresh
Top menu *Help* > *Cheatsheets*

# *Markdown wise
Markdown is great!
You can do websites, word docs, PDFs, books, etc.

## Trick - Add a code chunk
*Shortcut* `Cmd / Ctrl + Alt + I`

## Tip - Check code chunk options
I never remember so here is a cheat sheet^[[https://raw.githubusercontent.com/rstudio/cheatsheets/master/rmarkdown-2.0.pdf](https://raw.githubusercontent.com/rstudio/cheatsheets/master/rmarkdown-2.0.pdf)]

## Tip - Rmarkdown Theme Gallery
Use what is available ^[[https://www.datadreaming.org/post/r-markdown-theme-gallery/](https://www.datadreaming.org/post/r-markdown-theme-gallery/)] and
make sure you have pandoc installed, otherwise you might see some errors.

For example, if you are on Ubuntu focal ^[here is what I needed [https://packages.ubuntu.com/focal/pandoc](https://packages.ubuntu.com/focal/pandoc)].

This page and PDF are using the Tufte handout style ^[Tufte is a style that 
Edward Tufte uses in his books and handouts. Tufte's style is known for its extensive use of sidenotes, tight integration of graphics with text, and well-set typography. See Github repositories [tufte-latex](https://github.com/tufte-latex/tufte-latex), [tufte-css](https://github.com/edwardtufte/tufte-css) and its implementation
into the [**tufte** package](https://github.com/rstudio/tufte)].

For HTML output, use `tufte_html` in the YAML metadata at the beginning of 
an R Markdown document (see an example below).

```yaml
---
title: "An Example Using the Tufte Style"
author: "Paula Andrea Martinez"
output:
  tufte::tufte_handout: default
  tufte::tufte_html: default
---
```

`r newthought('Intermediate')`

# Shortcuts all

Alt+Shift+K

# *Manage your files

## Tip - check your files and folders
Lists your files and folders
```{r}
library(here)
dir(here("")) 
```

## Trick - get more information about files and folders
```{r}
fs::dir_info(here(""))
```
# *Writing code

## Tip - Naming things
Please watch Naming things from Jenny Bryan ^[[https://speakerdeck.com/jennybc/how-to-name-files](https://speakerdeck.com/jennybc/how-to-name-files)]

## Tip - replacing NAs
```{r}
library(tidyverse)
```

```{r}
msleep <- ggplot2::msleep
glimpse(msleep)
```
```{r}
msleep_nona <- msleep %>% 
  mutate(conservation = replace_na(conservation, "unknown"))

glimpse(msleep_nona)
```

## Tip - Selecting columns based on regex
```{r}
msleep %>%
  select(matches("wt")) %>%
  glimpse
```
## Tip - selecting by discrete number of categories
```{r}
msleep %>%
  select_if(~n_distinct(.) < 6)
```

```{r}
unique(msleep$vore)
```
## My favourite function count
```{r}
msleep %>% 
  count(vore)

```
## Tip - add count
```{r}

msleep %>%
  select(name:vore) %>%
  add_count(vore)
```

## Tip - get rid of extra characters in column names

```{r}
msleep_nona <- msleep_nona %>%
  select(1:4)

colnames(msleep_nona) <- c("Q1 Name", "Q2 sleeP total 1", "Q3 voore", "Q4 order")
colnames(msleep_nona)

msleep_nona %>%
  select_all(~str_replace(., "Q[0-9]+ ", "")) %>% 
  select_all(~str_replace_all(., " ", "_"))  
```

## Trick use Janitor to clean names
```{r}
library(janitor)

msleep_nona %>% 
  janitor::clean_names()

```

## Tip - Random selection of rows


```{r}
set.seed(123)

msleep %>%
  sample_frac(0.1)
```

## Trick - separate one column into two
```{r}
df <- data.frame(x = c("a:1", "a:2", "c:4", "d", NA))
(df)
```
```{r}
df %>% separate(x, c("key","value"), ":")
```
## Trick near()
```{r}
msleep %>% 
  select(name, sleep_total)
```
```{r}
sd(msleep$sleep_total)
```


```{r}
msleep %>% 
  select(name, sleep_total) %>% 
  filter(near(sleep_total, 17, tol = sd(sleep_total)))
```

## Tip - Use %in% instead of or
```{r}
y <- c("a", "a", "z", "y","b", "c")
```


```{r}
y == "a" | y == "b" | y == "c"  
```
is the same as
```{r}
y %in% c("a", "b", "c")
```



```{r}
msleep %>% 
  select(order, name, sleep_total) %>% 
  filter(order %in% c("Didelphimorphia", "Diprotodontia"))
```

## Trick - Use %in% instead of or

```{r}
msleep %>% 
  select(order, name, sleep_total) %>% 
  filter(order %in% str_subset(order, "Di"))

# add negative for second example
```
## Tip - vars all_vars
```{r}
msleep %>%  
  select(name, sleep_total:sleep_cycle) %>% 
  filter_at(vars(sleep_total, sleep_rem), all_vars(. > 5))
```

## Tip - beautiful plots
[https://twitter.com/wearerladies/status/1280859374668283906](https://twitter.com/wearerladies/status/1280859374668283906)

`r newthought('Advanced')`

# *Writing code

## Tip - Avoid dots in names
Believe me, or watch Jim Hester video ^[[https://www.youtube.com/watch?v=IoWDQ6rx6yA](https://www.youtube.com/watch?v=IoWDQ6rx6yA)]

# *Working with BIG files
Use when data is larger than 1 Giga

## Tip - Data table package
```{r eval=FALSE}
install.packages("data.table")
```

use the `fread()` function to read in big files

## Tip - Use fread from data.table and filter with grep
This is amazing!

```{r eval=FALSE}
data <- data.table::fread("grep -w File ~/data/someHUGEfile.csv")
```

pre-filter with the grep command ;)

## Trick - Use the pipe from base and filter

```{r eval=FALSE, echo=FALSE}
normal_data <- read.csv(pipe("grep -w File ~/data/somefile.csv"))
```


## Trick - Compress files on the fly!
```{r eval=FALSE}
write.csv(data, gzfile("data/bigdata.gz"))
file <- read.csv(gzfile("data/bigdata.gz"))
```

## Tip - Read or write big outputs
Functions in order of speed
```{r eval=FALSE}
read.csv()
readr::read_csv()
vroom::vroom_read()
```

## Tip - Read or Write to compressed formats
Compress directly into *gz, 7z, zst*. 
We know that `write.csv` from base does a pretty good job for most smallish files.
Try `readr::write_csv` and you will see the a 50% improvement. 
But, what surpasses all is `vroom::vroom_write` ~ 15 X faster than `write.csv`.

## Tip - benchmark
When things are working the next step is optimise!
You can check yourself with the function `bench_time` from the {bench} package.

# *Workflows and Reproducibility

## Tip - Use workflowr
Demonstration of a {workflowr} website ^[[https://www.youtube.com/watch?v=O1wv94sZfvE](https://www.youtube.com/watch?v=O1wv94sZfvE)]
Milestones, versions, all in one place.


## Tip - learn drake
If you are building analysis code that is likely to grow use drake
The {drake} package records file inter dependencies in your analysis. When files
are changed, {drake} only re-runs the parts that need to be re-run. This saves 
time and reduces errors ^[Learn from Matt Dray 
[https://www.rostrum.blog/2019/07/23/can-drake-rap/](https://www.rostrum.blog/2019/07/23/can-drake-rap/)
and from Amanda Dobbyn Rladies Chicago
[https://aedobbyn.github.io/nyc-fires/index.html#1]([https://aedobbyn.github.io/nyc-fires/index.html#1)].

`r newthought('Keep learning')`

Resources used to provide you with this collection of tips and tricks
  
* Suzan's tidyverse tricks [https://suzan.rbind.io/categories/tutorial/](https://suzan.rbind.io/categories/tutorial/)
* Sean Lopp's posts [https://rviews.rstudio.com/categories/tips-and-tricks/](https://rviews.rstudio.com/categories/tips-and-tricks/)
* Sean Lopp's video NYC RStudio Conference [https://www.youtube.com/watch?v=kuSQgswZdr8](https://www.youtube.com/watch?v=kuSQgswZdr8)
* Jim Hester's video Pipe Connections [https://www.youtube.com/watch?v=RYhwZW6ofbI](https://www.youtube.com/watch?v=RYhwZW6ofbI)
* tidyr reference docs [https://tidyr.tidyverse.org/reference](https://tidyr.tidyverse.org/reference)

*Author: Paula Andrea Martinez*
`r Sys.Date()`

```{r bib, include=FALSE}
# create a bib file for the R packages used in this document
knitr::write_bib(c('base', 'rmarkdown'), file = 'skeleton.bib')
```
